Do professional chess players draw too often?

### 2022-03-08 paper/paper.tex
adding rewrite

\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
    pdftitle={Overleaf Example},
    pdfpagemode=FullScreen,
}

@@

\section{The game}

In this paper we make a number of assumptions about each chess game observed.
First, we assume that each player plays at the strength of their respective Elo rating, or
in other words, plays their best. We make no attempt to model player strength directly or
measure the variance in their performance.
Instead, we want to know how aggressive or how passive they are playing.

Put more formally, given two players $W$ and $B$, we want to know
$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$, where $p_{\text{outcome}}$ is
the probability of $W$ (white) winning, drawing, and losing respectively.

@@

Given the ratings $R_A$ and $R_B$ of two players $A$ and $B$ respectively, the Elo rating system functions as follows.
\begin{align*}
    E_A = \frac{1}{1+10^{(R_B-R_A)/400}} \\
    \text{where } E_A = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}
\end{align*}
It follows from this that $E_A+E_B=1$, unsurprisingly,

\subsection{Player strategies and equilibrium}

The Elo rating system makes an implicit assumption that draw frequency does not directly affect
the expected score $E$.
There are infinite solutions to $E = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}$.
So, to simplify chess for our purposes, we assume that each player plays
according to their rating, and the choice they must make is how aggressive or passive to play.

If we assume, for instance, that both players $W$ and $B$, are playing at maximal aggression, then $E_W=p_{\text{win}}$ and $E_B=p_{\text{lose}}$. That is, $p_{\text{draw}}=0$.
But if both players are not maximally aggressive, the probability of a draw increases.
To write out all the strategies, we could define an index $a_W, a_B$ for each player that indicates their aggression,
and then rebalance
$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$ accordingly, but enumerating the strategies formally turns out not to matter.
This is because, under Elo assumptions, $E_W$ and $E_B$ are constant, regardless of draw probability.
So while players can change the probability of drawing depending on their strategies, they
cannot change their expected score.
% below is not correct, fix the math
% Both players can contribute to this probability, so $p_{\text{draw}}=p_{\text{draw}_W}+p_{\text{draw}_B}$,
% where $p_{\text{draw}_W}$ represents $W$'s passive play or propensity to draw.
% Under Elo assumptions, without changing their expected score, a player $A$ can choose to reduce
% $p_{\text{win}_A}$ by $f$ to increase $p_{\text{draw}_A}$ by $2f$ (and therefore increase $p_{\text{draw}}$ as well).
%
% Because this adjustment can result in $p_{\text{draw}}>1$, we limit each player's contribution to $0.5$, that is
% $p_{\text{draw}_A}<0.5$.
% This allows the total probability of a draw $p_{\text{draw}}$ to be $1$, which while unrealistic

@@

Our models will estimate these probabilities $p$ in order to determine whether these draw probabilities are in accordance with Elo's system, that is, reflective of player ability and playstyle,
or if they result from ``irrational play.''

The prediction, as described by Polgar \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{here}, is that real-world circumstances encourage drawing at the expense of good chess.
As a result of chess being played sub-optimally, we would expect that as a result of a player playing for a draw more aggressively,
they sacrifice some results. In other words, if you play to minimize variance, you decrease your expected score.

\section{The models}
The dataset used was Caissabase, a free database of over 5M professional chess games.
Simul games, Chess960 games, and consultation (team) games were removed.
Games played by players with less than 10 games were also removed.

(I will make better citations and give more background and explanations of these models by the final draft I promise. I am rushing at the moment.)

\subsection{Player model}
For modelling player behavior, a factor model was used, inspired by David Smerdon's
\href{https://www.davidsmerdon.com/?p=2168}{Fighting Chess Index}.
Variables computed for each player were
\begin{itemize}
    \item total games
    \item average rating
    \item average absolute difference between player ratings
    \item average number of draws
    \item average number of draws in 30 moves or less
    \item average number of draws in 30 moves or less when player was White
    \item average number of moves in drawn games
\end{itemize}
See Section 12.2.4 in Bishop's \textit{Pattern Recognition and Machine Learning} for a detailed description of this factor analysis model.

Smerdon uses PCA to form his Fighting Chess Index, but this method has its flaws. As described in Bishop,
PCA can be formulated as a maximum likelihood solution of a latent variable model.
But PCA does not treat the observed variables as conditionally independent given the latent variables.
So if we use principal components to measure ``aggression'' or ``fighting spirit,'' we are not treating
average rating (a measure of player strength) and irrational draw behavior as independent.
This is problem since they are correlated.
More serious (stronger) players have more on the line, and so will generally play for draws more often.
But as we want to measure ``fighting spirit'' for individual player independent of their strength, we do not want this dependence.

In Smerdon's defense, PCA's components do lend themselves to the construction of his index (whereas the factor model used here does not), but this does not change the problem of the underlying assumption.

The factor model used here does treat observations as conditionally independent of the latent variables.
This is tremendously useful, since we want to retain measures of player strength while independently encoding
draw behavior, especially as inputs to the following model.

\subsection{Game model}
(Again I will provide more detail wrt inference, etc in final, in a rush to the deadline as a result of repeated disasters)

For each game, we modelled $p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$
using the following generative model, where $l_w, l_b$ are the latent variables
 computed from the player model for the corresponding players.
\begin{align*}
    p = A l_w + B l_b + \epsilon \\
    \epsilon \sim \mathcal{N}(\mathbf{0}, \sigma\mathbf{I}) \\
    \sigma\sim \Gamma(\alpha=1,\beta=2)
\end{align*}

\section{Results}

\end{document}

>>>>

\documentclass{article}
\usepackage{amsmath}
\usepackage{hyperref}
\usepackage{graphicx}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    filecolor=magenta,
    urlcolor=blue,
}
\usepackage[parfill]{parskip}
\usepackage{bm}

\newcommand{\pw}{p_{\text{win}}}
\newcommand{\pd}{p_{\text{draw}}}
\newcommand{\pl}{p_{\text{lose}}}
\newcommand{\Rw}{R_{W_{\text{adj}}}}
\newcommand{\Rb}{R_{B_{\text{adj}}}}

@@

\section{Optimal chess}

Given two players $W$ and $B$, we want to know
$p=(p_{\text{win}}, p_{\text{draw}}, p_{\text{lose}})$, where $p_{\text{outcome}}$ is
the probability of $W$ (white) winning, drawing, and losing respectively.

@@

Given the ratings $R_A$ and $R_B$ of two players $A$ and $B$ respectively, the Elo rating system predicts game outcomes as follows.
\begin{gather*}
    Q_A = 10^{R_A/400}, Q_B = 10^{R_B/400} \\
    E_A = \frac{Q_A}{Q_A+Q_B} \\
    \text{where } E_A = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}
\end{gather*}
It follows from this that $E_A+E_B=1$.

\subsection{Equilibria under Elo assumptions}

The Elo rating system makes an implicit assumption that draw frequency does not directly affect
the expected score $E$, as it does not define $p_{\text{win}}$ directly.

There are infinite solutions to $E = p_{\text{win}} + \frac{1}{2} p_{\text{draw}}$.
Under Elo assumptions players can play solid passive chess and solid aggressive chess,
changing the probabilities of winning and drawing but importantly \textit{not changing}
the total expected value $E$.

@@

But if players play an overly passive style to achieve more draws,
they decrease their total expected value $E$ to increase $p_{\text{draw}}$.
This is the sort of behavior we are interested in, whether players play
too passively or draw too often \textit{at the expense} of playing optimal chess.

This is Polgar's prediction, as described \href{https://chessdailynews.com/the-700-lbs-gorilla-issue-to-draw-or-not-to-draw/}{here}; i.e., real-world circumstances encourage drawing over playing optimally.
As a result of chess being played sub-optimally, we would expect that as a result of a player playing for a draw more aggressively,
they sacrifice some results. In other words, if you play to minimize variance, you decrease your expected score.
To describe this, we introduce a modified Elo system.

\section{(Potentially) suboptimal chess}

First we describe the framework for describe hyperpassive, suboptimal chess before discussing potential equilibria.

\subsection{Modified Elo game}

There are two players, $W, B$, white and black, with ratings $R_W,R_B$ respectively.

Each chooses how ``overly passive'' to play, that is, how much they want to
sacrifice expected value for increased draw percentage.

For both $W,B$, their strategies are to choose $$\alpha_W, \alpha_B \in [0,1]$$

Their payoffs, $\Pi_W,\Pi_B$, are represented by the following modified Elo equations.
Similar to the standard Elo equations above, they represent the expected score of each player,
but instead the expected score is scaled by $\alpha_W, \alpha_B$.
% , and the probability of winning is scaled by a sigmoid-weighted average of $\alpha_W,\alpha_B$.
\begin{gather*}
    % \sigma(x)=\frac{1}{1+\exp(-x)} \\
    Q_W = 10^{R_W/400}, Q_B = 10^{R_B/400} \\
    \Pi_W = \frac{\alpha_W Q_W}{\alpha_W Q_A+\alpha_B Q_B}
    % = \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
    % {\sigma(\alpha_W) + \sigma(\alpha_B)}
    % P(W\text{ Win}) + \frac{1}{2} P(\text{Draw})
    \\
    \Pi_B = \frac{\alpha_B Q_B}{\alpha_W Q_A+\alpha_B Q_B}
    % = \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
    % {\sigma(\alpha_W) + \sigma(\alpha_B)}
    % P(B\text{ Win}) + \frac{1}{2} P(\text{Draw})
\end{gather*}

If a player $A$ chooses $\alpha=1$, then they play optimally, that is
$\alpha_A Q_A = Q_A$.
If they choose $\alpha=0$ then they ``turtle,''
give up all chances of winning only to increase the likelihood of drawing.
Importantly, if $\alpha_W=\alpha_B$, that is if both players are playing
equally passively, then $\Pi_W=E_W$ and $\Pi_B=E_B$; their payoffs
remain the same as in the standard Elo system, as they are \textit{both}
playing equally suboptimally.

Importantly, scaling $Q_W,Q_B$ by $\alpha_W,\alpha_B$ is equivalent to changing
each player's effective rating by $R_W + 400\log_{10}(\alpha_W),R_B + 400\log_{10}(\alpha_B)$.
This fact is used in the code when estimating $\alpha_A, \alpha_B$.

This difference in rating is not the only effect that $\alpha_A,\alpha_B$ have,
since this modified Elo description deals only with suboptimal play and changing expected values;
the issue of changing draw probability will be addressed in the next section.

\subsection{Estimating outcome probabilities}
To estimate $p=(\pw,\pd,\pl)$, we initially use a multinomial regression as follows.
\begin{gather*}
    \bm{x} = (R_W, R_B)\\
    \pw = \frac{\exp(-\bm{\beta}_w \cdot \bm{x})}
    {1+\exp(-\bm{\beta}_w \cdot \bm{x}) + \exp(-\bm{\beta}_d \cdot \bm{x})} \\
    \pd = \frac{\exp(-\bm{\beta}_d \cdot \bm{x})}
    {1+\exp(-\bm{\beta}_w \cdot \bm{x}) + \exp(-\bm{\beta}_d \cdot \bm{x})} \\
    \pl = 1 - \pw - \pd
\end{gather*}
To account for $\alpha_A, \alpha_B$ changing draw probabilities, we make the following
modifications to $\pw,\pd,\pl$,
beginning with the rating adjustments as described in the previous section.
\begin{gather*}
    R_{W_{\text{adj}}} = R_W + 400\log_{10}(\alpha_W), R_{B_{\text{adj}}} = R_B + 400\log_{10}(\alpha_B) \\
    \bm{x}'=(\Rw, \Rb)\\
    \sigma(x) = \frac{1}{1+\exp(-x)} \\
    \gamma(\alpha_W, \alpha_B) = 1 - \frac{\sigma(\alpha_W)\alpha_W + \sigma(\alpha_B)\alpha_B}
    {\sigma(\alpha_W)+\sigma(\alpha_B)} \\
    p_{\text{win}_{\text{adj}}} = \frac{\exp(-\bm{\beta}_w \cdot \bm{x}')}
    {1+\exp(-\bm{\beta}_w \cdot \bm{x}') + \exp(-\bm{\beta}_d \cdot \bm{x}')} \\
    p_{\text{draw}_{\text{adj}}} = \frac{\exp(-\bm{\beta}_d \cdot \bm{x})}
    {1+\exp(-\bm{\beta}_w \cdot \bm{x}') + \exp(-\bm{\beta}_d \cdot \bm{x}')}
    + \gamma(\alpha_W,\alpha_B)\left( \pw - p_{\text{win}_{\text{adj}}} \right)\\
    p_{\text{lose}_{\text{adj}}} = 1- p_{\text{win}_{\text{adj}}} - p_{\text{draw}_{\text{adj}}}
\end{gather*}
The function $\gamma$ is a sigmoid-weighted average of the two $\alpha$ parameters.
It represents the fraction of draw probability that is gained by sacrificing win percentage.
In other words, it represents the decrease in variance that comes at the cost of decreasing
expected value.

If both players choose the same $\alpha$, then the weighted average
will come out to just $\alpha$, otherwise the average biases toward
the player playing more optimally (that is, the player with the greater $\alpha$).
That is, it is more punishing to the player playing more suboptimally passively; the
player playing worse does not receive a proportional increase in draw percentage.

\subsection{Defining $\alpha_W,\alpha_B$}
For modelling player behavior, the following feature variables were used, inspired by David Smerdon's
\href{https://www.davidsmerdon.com/?p=2168}{Fighting Chess Index}.
Variables computed for each player in the dataset were
\begin{itemize}
    \item total games
    \item average rating
    \item average absolute difference between player ratings
    \item average number of draws
    \item average number of draws in 30 moves or less
    \item average number of draws in 30 moves or less when player was White
    \item average number of moves in drawn games
\end{itemize}
Smerdon uses these features as inputs to PCA to compute an index of player aggression.
Here, we use these features to estimate $\alpha_W, \alpha_B$.

We add two vectors of weights $\bm{w}_W,\bm{w}_B$ to the model to represent
linear combinations of the above features, $\bm{f}_W,\bm{f}_B$ for white and black respectively, with
intercept terms $\delta_W,\delta_B$.
Importantly, all features above were $z$-scored.
\begin{gather*}
    \alpha_W=\delta_W + \bm{w}_W\cdot \bm{f}_W \\
    \alpha_B=\delta_B + \bm{w}_B\cdot \bm{f}_B
\end{gather*}

\subsection{Learning parameters}
There are no hyperparameters in the model. All parameters
$\bm{\beta}_d,\bm{\beta}_w, \delta_W, \delta_B,\bm{w}_W,\bm{w}_B$ are learned through maximum likelihood estimation
according to the computed multinomial distribution
$p=(p_{\text{win}_{\text{adj}}},p_{\text{draw}_{\text{adj}}},p_{\text{lose}_{\text{adj}}})$.

For more information see the GitHub repo.

\section{Results}
\subsection{Data}
The dataset used was \href{http://caissabase.co.uk/}{Caissabase}, a free database of 5.61 million professional chess games.
Simul games, Chess960 (a chess variant) games, and consultation (team) games were removed.
Games played by players with fewer than 10 games were also removed.
Games played by chess engines were also removed, as were games with invalid player names.

Player names were cleaned and standardized using Levenshtein distance fuzzy string matching,
see \href{https://github.com/seatgeek/thefuzz}{this Python package} for more details.

This left 4,695,527 games in the dataset.

\subsection{Model selection}
To test the statistical significance of all the additional parameters, a parsimonious model
was formulated that excludes $\alpha_W,\alpha_B$ and all additional parameters, instead only
taking $R_W,R_W$.

A likelihood ratio was computed followed by a chi-squared test, yielding a $p$-value of $<0.000001$.
This is not unexpected, since those parameters describe draw behavior, whereas ratings and the Elo system more generally, as discussed
below, do little to describe draw probabilities.

\subsection{Distributions of $\alpha_W, \alpha_B$}

Below are kernel density estimates of the densities of $\alpha_W,\alpha_B$.

\includegraphics[width=\textwidth]{./distributions.png}

Interestingly, the density at $\alpha_W=\alpha_B=1$ is almost $0$.
This might suggest that no players are playing perfectly optimally, but more likely than not,
this indicates one of the flaws of the Elo rating system.

It is clearly useful (and very effective) for ranking players, but does poorly when applied
to draw prediction. This is not relevant for its main purpose, but poses a challenge here.
Because the Elo expected value $$E_W=\pw + \frac{1}{2}\pd$$
neither defines $\pw$ nor $\pd$, instead assigning a ``payoff'' of $\frac{1}{2}$ to drawing,
it becomes difficult to predict draw behavior using it.
This is especially visible at high levels of play where draws are quite common;
35\% of games in the cleaned dataset were draws.
This flaw was also present in the
modified Elo expected value, which was more to
derive relative playing strength in the form of adjusted ratings, rather than directly compute draw probabilities.

So rather than looking at the individual distributions of $\alpha_W,\alpha_B$, it is more informative to examine
the distribution of their difference.

\includegraphics[width=\textwidth]{./diff.png}

This is almost zero-mean, with the average difference being $0.0015$ and the median difference being $0.0055$.
It should be expected that white plays less passively, since white wins 37\% of the time as compared
to black's 28\%
(in this dataset).
Interestingly the tails of the $\alpha_B$ distribution are quite long in comparison to the short tails of the $\alpha_W$ distribution.
This could again be due to black's native disadvantage encouraging more deviation and more strategies chosen
to either really attempt to win (higher $\alpha_B$) or attempt to draw at at all costs (lower $\alpha_W$).

\includegraphics[width=\textwidth]{./linear.png}

There is also a correlation between rating and $\alpha$, with an $R$-value of 0.5047, indicating that at higher ratings
there is less suboptimal play.
This is not totally unsurprising, since at the highest echelons of play this suboptimal playing behavior is much more easily
observed, socially discouraged, and more heavily punished by other players (due to the higher levels of play).
At slightly lower levels of play, from national masters to lower-rated international masters, though, we observe lower
values of $\alpha$.

\section{Conclusion}
These results relate back to the article/real-world situation in a couple of ways.
First, it shows that variance-lowering behavior at the cost of expected value does occur in chess.
That is, playing suboptimally for draws at higher levels of play occurs often.
It is difficult to identify, however, if this is as a result of random noise or variance in playing strength
on behalf of players, or if each choice of $\alpha_W, \alpha_B$ for each game is intentional.
It is also clear that this behavior, on average, decreases as the level of play increases, as predicted.

That being said, the majority of situations involve players playing similarly, that is choosing
$\alpha_W\approx \alpha_B$, so although some suboptimal hyperpassive play occurs, it does not occur
sufficiently to pose a threat to the integrity of the game.

The article mentions measures designed to increase errors and decrease the likelihood of draws.
These would likely be successful at the highest echelons of play, where games are sufficiently
scrutinized and play is so precise that suboptimal play essentially does not exist.
However, at slightly lower levels of play (still professional, but just among weaker players), the problem is not absolute precision.
Rather, (assuming the results support this), the problem is this variance-reducing behavior.
So from a design perspective, the solution would be to penalize this behavior.
Making draws worth $\frac{1}{3}$ of a point could discourage taking draws in tournaments, for instance.
Removing agreed draws (where players simply agree to a draw rather than playing it out) as a whole could also make this behavior more difficult.

% TODO BIB
Caissabase, Elo, Smerdon, The Fuzz, original article
\end{document}
